---
title: "Performance on real EWAS-data"
author: "Bai Ling"
date: "`r format(Sys.Date(), format = '%d %B %Y')`"
output: 
  html_document:
    theme: cerulean
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
options(stringsAsFactors = FALSE)
setwd("/Users/apple/Documents/GitHub/EWAS/")
```

## Contents
***
- [1. Run time for each method on real EWAS data](#runtime)
- [2. Signal numbers: ST vs BH](#STvsBH)
- [3. Overview of changes across methods and covariates](#overview)  
- [4. An example demonstrated the strong informativeness of covariate mean](#example)
- [5. The relationship between omibus p-value and signal changes](#relationship)
<br>
<br>
<br>

#### Library R packages and basic settings
```{r library prepare, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(reshape2)
library(circlize)
library(ComplexHeatmap)
library(RColorBrewer)
library(cowplot)
library(ggplotify)
library(ggpubr)
library(ggforce)
method_col <- c("CAMT" = "#FF0000", "AdaPT" = "#E69F00", "IHW" = "#56B4E9", 
                "FDRreg" = "#009E73", "BL" = "#0072B2")
measure_col <- c("BH"="#1B9E77", "ST"="#D95F02", "sd.b"="#7570B3", "sd.m"="#E7298A", "mean"="#A6CEE3", 
                 "mad"="#1F78B4", "dip"="#B2DF8A", "precision"="#FFFF00", "icc.b"="#FB9A99", 
                 "icc.m"="#E31A1C", "refgene.pos"="#FDBF6F", "chr"="#A6761D", "cpg.loc"="#CAB2D6",
                 "dhs"="#949494", "direction"="#7FFFD4", "probe.type" ="#FFE4E1")
```

### 1. Run time for each method on real EWAS data {#runtime}
***
```{r run time, fig.height=4, fig.width=6}
runtime <- readRDS("Data/runtime.RDs")
runtime <- mutate(runtime, log_runtime = log10(Run_Time)) %>%
  group_by(measure, method) %>%
  summarise(n = n(),
            mean = mean(log_runtime),
            sd = sd(log_runtime),
            se = sd / sqrt(n))

selected_methods <- c("AdaPT", "BL", "CAMT", "FDRreg", "IHW")
selected_measure <- c("dip", "icc.b", "icc.m", "mad", "mean", "precision", "sd.b", "sd.m",
                         "dhs", "direction", "probe.type", "cpg.loc", "refgene.pos", "chr")

fs6 <- runtime %>%
  filter(method %in% selected_methods, measure %in% selected_measure) %>%
  ggplot(aes(x = measure, y = mean, group = method, color = method)) +
  geom_line(lwd = 0.4) +
  geom_point(size = 0.5) +
  geom_errorbar(aes(ymin = mean-1.96*se, ymax = mean+1.96*se), width = 0.15, lwd = 0.35) +
  ylab("log10 Runtime(s)") + xlab("") +
  theme_bw() +
  theme(title = element_text(size = 10), 
         text = element_text(size = 8), 
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8), 
         axis.text.y = element_text(size = 8),
         legend.position = "right", 
         legend.justification = "center",
         legend.title = element_blank()) +
   scale_x_discrete(limits = selected_measure) +
   scale_color_manual(values = method_col)

fs6
```

```{r, include=FALSE}
save_plot("Figures/FigureS6.pdf", fs6)
```

### 2. Signal numbers: ST vs BH {#STvsBH}
*** 
For each data set, the number of ST detected significant differential CpGs is always no less than that of BH. So the result of ST is used as reference in the comparison with other covariate-adaptive FDR control methods.
```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
reference_discovery <- read.csv("Data/reference_discovery.csv", header = TRUE)
fs7 <- reference_discovery %>%
  ggplot(aes(x = BH + 1, y = ST + 1)) +
  geom_point() +
  geom_abline(slope = 1, color = "red") +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  xlab("BH") + ylab("ST") +
  theme(title = element_text(size = 10), 
         text = element_text(size = 8))
fs7
```

```{r, include=FALSE}
save_plot("Figures/FigureS7.pdf", fs7)
```

### 2. Overview of changes across methods and covariates{#overview}
***
#### 2.1 Heatmap: Performance evaluation by signal changes
```{r heatmp, message=FALSE, fig.height=6, fig.width=7.5}
discovery_number <- read.csv("Data/discovery_number.csv", header = TRUE)
covariate_order <- c("mean", "sd.b", "mad", "precision", "sd.m", "icc.m", "cpg.loc",
                     "direction", "icc.b", "refgene.pos", "probe.type", "dip", "dhs", "chr")

plot_data <- left_join(discovery_number, reference_discovery, by = "dataset") %>%
  filter(! measure %in% c("icc.b", "icc.m")) %>%
  arrange(pi0) %>%
  mutate(logFC = log2((discovery + 1)/(ST + 1)))

# prepare the matrix for heatmap
plot_data$measure <- factor(plot_data$measure, levels = covariate_order)
plot_data$dataset <- factor(plot_data$dataset, levels = unique(plot_data$dataset))
heatmap_data <- dcast(plot_data, dataset ~ method + measure, value.var = "logFC")
rownames(heatmap_data) <- heatmap_data$dataset
heatmap_data$dataset <- NULL
heatmap_data <- as.matrix(heatmap_data)

# prepare row and column annotation
annotation_row <- plot_data %>%
  select(dataset, ST, pi0)
annotation_row <- annotation_row[! duplicated.data.frame(annotation_row), ]

# top annotation: ranks of discovered signal number recovered by different covariates within each dataset
top_anno <- plot_data %>%
  group_by(dataset, method) %>%
  mutate(rank = rank(discovery, ties.method = "average")) %>%
  dcast(dataset ~ method + measure, value.var = "rank")
top_anno$dataset <- NULL

# heatmap
col_fun <- colorRamp2(c(-1, -0.5, 0, 0.5, 1), c("royalblue", "blue", "white", "red", "red3"), space = "RGB")
ST_color <- colorRamp2(c(0, 200, 1000, 10000), c("#DEEBF7", "#C6DBEF", "#6BAED6", "#08306B"))
pi0_color <- colorRamp2(c(0.6, 0.9, 0.95, 1), c("#F7FCF5", "#A1D99B", "#41AB5D", "#00441B"))

f3 <- Heatmap(heatmap_data,
              name = "logFC",
              rect_gp = gpar(col = "grey", lwd = 0.05),
              cluster_rows = FALSE,
              cluster_columns = FALSE,
              col = col_fun, 
              left_annotation = rowAnnotation(
                ST = annotation_row$ST,
                pi0 = annotation_row$pi0,
                col = list(ST = ST_color,
                           pi0 = pi0_color),
                simple_anno_size = unit(0.25, "cm"),
                annotation_name_gp = gpar(fontsize = 8),
                annotation_legend_param = list(title_gp = gpar(fontsize = 9),
                                               labels_gp = gpar(fontsize = 7))),
              top_annotation = columnAnnotation(
                rank = anno_boxplot(top_anno, outline = FALSE, 
                gp = gpar(fill = measure_col[covariate_order[-grep("icc", covariate_order)]]))),
              row_names_gp = gpar(fontsize = 5), 
              column_names_gp = gpar(fontsize = 8),
              column_split = rep(c("AdaPT", "BL", "CAMT", "FDRreg", "IHW"), each = 12),
              column_gap = unit(0, "mm"),
              column_labels = unlist(lapply(str_split(colnames(heatmap_data), pattern = "_"),
                                                 function(x)x[[2]])),
             heatmap_legend_param = list(title_gp = gpar(fontsize = 9),
                                         labels_gp = gpar(fontsize = 7)))

f3
```

```{r, include=FALSE}
pdf("Figures/Figure3.pdf", width = 6.69, height = 6)
draw(f3)
dev.off()
```

#### 2.2 Boxplot of logFC corresponding to the heatmap
```{r, fig.height=4, fig.width=6}
# Method AdaPT's performance is very different from other methods, so an alternative range is set for it.

p1 <- plot_data %>%
  filter(method == "AdaPT") %>%
  ggplot(aes(x = measure, y = logFC)) +
  geom_boxplot(aes(fill = measure), outlier.shape = NA, fatten = 1) +
  scale_fill_manual(values = measure_col) +
  coord_cartesian(ylim = c(-6, 6)) +
  facet_wrap(~ method) +
  theme_bw() +
  xlab("") +
  theme(title = element_text(size = 10),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

p2 <- plot_data %>%
  filter(method != "AdaPT") %>%
  ggplot(aes(x = measure, y = logFC)) +
  geom_boxplot(aes(fill = measure), outlier.shape = NA, fatten = 1) +
  scale_fill_manual(values = measure_col) +
  coord_cartesian(ylim = c(-2, 2)) +
  facet_wrap(~ method, nrow = 1, scales = "free_y") +
  theme_bw() +
  xlab("") + ylab("") +
  theme(title = element_text(size = 10),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "right",
        legend.justification = "center")

fs8_1 <- plot_grid(p1, p2, rel_widths = c(1, 4))
fs8_1
```
<br>
<br>

#### 2.3 Boxplot of logFC for each Method with the best informative covariate
```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
fs8_2 <- left_join(discovery_number, reference_discovery, by = "dataset") %>%
  mutate(logFC = log2((discovery + 1)/(ST + 1))) %>%
  group_by(dataset, method) %>%
  summarize(logFCmax = max(logFC)) %>%
  ggplot(aes(x = method, y = logFCmax, color = method)) +
  geom_boxplot(outlier.shape = NA, fatten = 1, lwd = 0.3) +
  geom_sina(size = 1, alpha = 0.5, shape = 16) +
  coord_cartesian(ylim = c(-2, 3)) +
  scale_color_manual(values = method_col) +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        title = element_text(size = 10)) +
  xlab("") + ylab("logFC")

fs8_2
```
<br>
<br>
```{r}
fs8 <- plot_grid(fs8_1, fs8_2, labels = c("A", "B"), ncol = 1, label_size = 18)
```

```{r, include=FALSE}
save_plot("Figures/FigureS8.pdf", fs8, base_width = 6.69, base_height = 4)
```

### 4. An example demonstrated the strong informativeness of covariate mean {#example}
***
```{r, fig.width=7, fig.height=2.5}
EWAS33 <- readRDS("Data/dataset/EWAS33.RDS")
candidate_covariates <- c("sd.b", "mad", "mean", "precision")

EWAS33_data <- data.frame()
for(s in candidate_covariates){
  df <- data.frame(covariate_value = EWAS33[[s]],
                   p_value = EWAS33$P.value)
  df$covariate <- s
  f <- ecdf(df$covariate_value)
  df$percentile_covariate <- f(df$covariate_value)
  EWAS33_data <- rbind(EWAS33_data, df)
}

fs9 <- ggplot(data = EWAS33_data, mapping = aes(x = percentile_covariate, y = -log10(p_value))) +
  geom_hex(bins = 100) + 
  facet_wrap(~covariate, nrow = 1) + 
  theme_bw() +
  theme(legend.key.size = unit(0.3, "cm"))
fs9
```

```{r, include=FALSE}
save_plot("Figures/FigureS9.pdf", plot = fs9, base_width = 7, base_height = 2.5)
```

### 5. Signal number changes vs omnibus test p-value {#relationship}
Three groups ("p<=0.001", "0.001<p<=0.01", "p>0.01") are classified according to omnibus test p-value. We explored the signal number changes accross the three groups.
```{r, fig.width=6.69, fig.height=3.5, message=FALSE, warning=FALSE}
omnibus_result <- read.csv("Data/omnibus.result.csv", header = TRUE)

# Method AdaPT's performance is very different from other methods, so an alternative range is set for it.

p1 <- left_join(plot_data, omnibus_result) %>%
  filter(method == "AdaPT") %>%
  mutate(group = factor(ifelse(p.value <= 0.001, "p<=0.001", ifelse(p.value <= 0.01, "0.001<p<=0.01", "p>0.01")),
                        levels = c("p<=0.001", "0.001<p<=0.01", "p>0.01"))) %>%
  ggplot(aes(x = group, y = logFC)) +
  geom_boxplot(aes(fill = group), outlier.shape = NA, fatten = 1) +
  stat_summary(fun.y = mean, geom = "point", shape = 18, color = "black") +
  coord_cartesian(ylim = c(-6, 6)) +
  facet_wrap(~method, nrow = 1, scales = "free") +
  theme_bw() +
  theme(title = element_text(size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

p2 <- left_join(plot_data, omnibus_result) %>%
  filter(method != "AdaPT") %>%
  mutate(group = factor(ifelse(p.value <= 0.001, "p<=0.001", ifelse(p.value <= 0.01, "0.001<p<=0.01", "p>0.01")),
                        levels = c("p<=0.001", "0.001<p<=0.01", "p>0.01"))) %>%
  ggplot(aes(x = group, y = logFC)) +
  geom_boxplot(aes(fill = group), outlier.shape = NA, fatten = 1) +
  stat_summary(fun.y = mean, geom = "point", shape = 18, color = "black") +
  coord_cartesian(ylim = c(-2, 2)) +
  facet_wrap(~method, nrow = 1, scales = "free") +
  theme_bw() +
  theme(title = element_blank(),
        axis.text = element_text(size = 8),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right")

fs10 <- plot_grid(p1, p2, rel_widths = c(1, 4.5))
fs10
```

```{r, include=FALSE}
save_plot("Figures/FigureS10.pdf", fs10, base_width = 8)
```

