---
title: "Performance on real EWAS-data"
author: "Bai Ling"
date: "`r format(Sys.Date(), format = '%d %B %Y')`"
output: 
  html_document:
    theme: cerulean
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
options(stringsAsFactors = FALSE)
setwd("/Users/apple/Documents/GitHub/EWAS/")
```

## Contents
***
- [1. Run time for each method on real EWAS data](#runtime)
- [2. Signal numbers: ST vs BH](#STvsBH)
- [2. Overview of changes across methods and covariates](#overview)  
- [3. An example demonstrated the strong informativeness of covariate mean](#example)
<br>
<br>
<br>

#### Library R packages and basic settings
```{r library prepare, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(reshape2)
library(circlize)
library(ComplexHeatmap)
library(RColorBrewer)
library(cowplot)
library(ggplotify)
library(ggpubr)
library(ggforce)
method_col <- c("CAMT" = "#FF0000", "AdaPT" = "#E69F00", "IHW" = "#56B4E9", 
                "FDRreg" = "#009E73", "BL" = "#0072B2")
measure_col <- c("BH"="#1B9E77", "ST"="#D95F02", "sd.b"="#7570B3", "sd.m"="#E7298A", "mean"="#A6CEE3", 
                 "mad"="#1F78B4", "dip"="#B2DF8A", "precision"="#FFFF00", "icc.b"="#FB9A99", 
                 "icc.m"="#E31A1C", "refgene.pos"="#FDBF6F", "chr"="#A6761D", "cpg.loc"="#CAB2D6",
                 "dhs"="#949494", "direction"="#7FFFD4", "probe.type" ="#FFE4E1")
```

### 1. Run time for each method on real EWAS data {#runtime}
***
```{r run time}
runtime <- readRDS("Data/runtime.RDs")
runtime <- mutate(runtime, log_runtime = log10(Run_Time)) %>%
  group_by(measure, method) %>%
  summarise(n = n(),
            mean = mean(log_runtime),
            sd = sd(log_runtime),
            se = sd / sqrt(n))

selected_methods <- c("AdaPT", "BL", "CAMT", "FDRreg", "IHW")
selected_measure <- c("dip", "icc.b", "icc.m", "mad", "mean", "precision", "sd.b", "sd.m",
                         "dhs", "direction", "probe.type", "cpg.loc", "refgene.pos", "chr")

panel_A <- runtime %>%
  filter(method %in% selected_methods, measure %in% selected_measure) %>%
  ggplot(aes(x = measure, y = mean, group = method, color = method)) +
  geom_line(lwd = 0.3) +
  geom_point(size = 0.025) +
  geom_errorbar(aes(ymin = mean-1.96*se, ymax = mean+1.96*se), width = 0.2, lwd = 0.2) +
  ylab("log10(seconds)") + xlab("") +
  theme(title = element_text(size = 10), 
         text = element_text(size = 8), 
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8), 
         axis.text.y = element_text(size = 8),
         legend.position = "right", 
         legend.justification = "center",
         legend.title = element_blank()) +
   scale_x_discrete(limits = selected_measure) +
   scale_color_manual(values = method_col)

panel_A
```

### 2. Signal numbers: ST vs BH {#STvsBH}
***
```{r, message=FALSE, warning=FALSE}
reference_discovery <- read.csv("Data/reference_discovery.csv", header = TRUE)
fs5 <- reference_discovery %>%
  ggplot(aes(x = BH + 1, y = ST + 1)) +
  geom_point() +
  geom_abline(slope = 1, color = "red") +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  xlab("BH") + ylab("ST")
fs5
```
```{r, include=FALSE}
save_plot("Figures/FigureS5.pdf", fs5)
```



### 2. Overview of changes across methods and covariates{#overview}
***
#### 2.1 heatmap
```{r heatmp, message=FALSE, fig.height=6, fig.width=7.5}
discovery_number <- read.csv("Data/discovery_number.csv", header = TRUE)
covariate_order <- c("mean", "sd.b", "mad", "precision", "sd.m", "icc.m", "cpg.loc",
                     "direction", "icc.b", "refgene.pos", "probe.type", "dip", "dhs", "chr")

plot_data <- left_join(discovery_number, reference_discovery, by = "dataset") %>%
  filter(! measure %in% c("icc.b", "icc.m")) %>%
  arrange(pi0) %>%
  mutate(logFC = log2((discovery + 1)/(ST + 1)))

# prepare the matrix for heatmap
plot_data$measure <- factor(plot_data$measure, levels = covariate_order)
plot_data$dataset <- factor(plot_data$dataset, levels = unique(plot_data$dataset))
heatmap_data <- dcast(plot_data, dataset ~ method + measure, value.var = "logFC")
rownames(heatmap_data) <- heatmap_data$dataset
heatmap_data$dataset <- NULL
heatmap_data <- as.matrix(heatmap_data)

# prepare row and column annotation
annotation_row <- plot_data %>%
  select(dataset, ST, pi0)
annotation_row <- annotation_row[! duplicated.data.frame(annotation_row), ]

# top annotation: ranks of discovered signal number recovered by different covariates within each dataset
top_anno <- plot_data %>%
  group_by(dataset, method) %>%
  mutate(rank = rank(discovery, ties.method = "average")) %>%
  dcast(dataset ~ method + measure, value.var = "rank")
top_anno$dataset <- NULL

# heatmap
col_fun <- colorRamp2(c(-1, -0.5, 0, 0.5, 1), c("royalblue", "blue", "white", "red", "red3"), space = "RGB")
ST_color <- colorRamp2(c(0, 200, 1000, 10000), c("#DEEBF7", "#C6DBEF", "#6BAED6", "#08306B"))
pi0_color <- colorRamp2(c(0.6, 0.9, 0.95, 1), c("#F7FCF5", "#A1D99B", "#41AB5D", "#00441B"))

panel_B <- Heatmap(heatmap_data,
                   name = "logFC",
                   rect_gp = gpar(col = "grey", lwd = 0.05),
                   cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   col = col_fun, 
                   
                   left_annotation = rowAnnotation(
                     ST = annotation_row$ST,
                     pi0 = annotation_row$pi0,
                     col = list(ST = ST_color,
                                pi0 = pi0_color),
                     simple_anno_size = unit(0.25, "cm"),
                     annotation_name_gp = gpar(fontsize = 8),
                     annotation_legend_param = list(title_gp = gpar(fontsize = 9),
                     labels_gp = gpar(fontsize = 7))),
                   
                   top_annotation = columnAnnotation(
                     rank = anno_boxplot(top_anno, outline = FALSE, 
                    gp = gpar(fill = measure_col[covariate_order[-grep("icc", covariate_order)]]))),
                   
                   row_names_gp = gpar(fontsize = 5), 
                   column_names_gp = gpar(fontsize = 8),
                   column_split = rep(c("AdaPT", "BL", "CAMT", "FDRreg", "IHW"), each = 12),
                   column_gap = unit(0, "mm"),
                   column_labels = unlist(lapply(str_split(colnames(heatmap_data), pattern = "_"),
                                                 function(x)x[[2]])),
                   heatmap_legend_param = list(title_gp = gpar(fontsize = 9), 
                                               labels_gp = gpar(fontsize = 7)))

panel_B
panel_B <- grid.grabExpr(draw(panel_B))
```
<br>
<br>

#### 2.2 Method with the best informative covariate
```{r, message=FALSE, warning=FALSE}
panel_C <- left_join(discovery_number, reference_discovery, by = "dataset") %>%
  mutate(logFC = log2((discovery + 1)/(ST + 1))) %>%
  group_by(dataset, method) %>%
  summarize(logFCmax = max(logFC)) %>%
  ggplot(aes(x = method, y = logFCmax, color = method)) +
  geom_boxplot(outlier.shape = NA, fatten = 1, lwd = 0.25) +
  geom_sina(size = 1, alpha = 0.2, shape = 16) +
  scale_color_manual(values = method_col) +
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        title = element_text(size = 10)) +
  xlab("") + ylab("logFC")

panel_C
```
<br>
<br>

#### 2.3 Boxplot of logFC corresponding to the heatmap
```{r}
fs6_1 <- plot_data %>%
  ggplot(aes(x = measure, y = logFC)) +
  geom_boxplot(aes(fill = measure), outlier.alpha = 0.1, outlier.size = 1, 
               outlier.shape = 16, fatten = 1) +
  scale_fill_manual(values = measure_col) +
  facet_wrap(~method, nrow = 1, scales = "free_x") +
  theme_bw() +
  xlab("") +
  theme(title = element_text(size = 10),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.key.size = unit(0.3, "cm"),
        legend.position = "top",
        legend.justification = "center")
fs6_1
```
<br>
<br>

#### 2.4 Signal number changes vs omnibus test p-value
```{r, fig.width=6.69, fig.height=3.5, message=FALSE, warning=FALSE}
omnibus_result <- read.csv("Data/omnibus.result.csv", header = TRUE)

fs6_2 <- left_join(plot_data, omnibus_result) %>%
  mutate(group = ifelse(p.value <= 0.001, "p<=0.001", ifelse(p.value <= 0.01, "p<=0.01", ifelse(p.value <= 0.05, "p<=0.05", "p>0.05")))) %>%
  ggplot(aes(x = group, y = logFC)) +
  geom_boxplot(aes(fill = group), outlier.alpha = 0.1, outlier.shape = 16, 
               outlier.size = 1, fatten = 1) +
  facet_wrap(~method, nrow = 1, scales = "free") +
  theme_bw() +
  theme(legend.position = "top") +
  theme(title = element_text(size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())
fs6_2


# fs6_3 <- left_join(plot_data, omnibus_result) %>%
#   mutate(group = ifelse(p.value <= 0.001, "p<=0.001", ifelse(p.value <= 0.01, "p<=0.01", "p>0.01"))) %>%
#   ggplot(aes(x = group, y = logFC)) +
#   geom_boxplot(aes(fill = group), outlier.alpha = 0.1, outlier.shape = 16, 
#                outlier.size = 1, fatten = 1) +
#   facet_wrap(~method, nrow = 1, scales = "free") +
#   theme_bw() +
#   theme(legend.position = "top") +
#   theme(title = element_text(size = 10),
#         axis.text = element_text(size = 8),
#         axis.text.x = element_blank(),
#         axis.title.x = element_blank(),
#         axis.ticks.x = element_blank())
# 
# fs6_3

```
<br>
<br>
```{r}
fs6 <- plot_grid(fs6_1, fs6_2, align = "v", axis = "l", labels = "AUTO", ncol = 1,
                 rel_heights = c(1.4, 1), label_size = 18)
fs6
```

```{r, include=FALSE}
save_plot("Figures/FigureS6.pdf", plot=fs6, base_width = 9, base_height = 6)
```



#### Organize figures into Figure 5
```{r, fig.width=7, fig.height=8}
top_row <- plot_grid(panel_A, panel_C,
                     align = "h", axis = "b", 
                     labels = c("A", "C"), label_size = 18,
                     rel_widths = c(1.2, 1),
                     nrow = 1)
f5 <- plot_grid(top_row, panel_B, labels = c("", "B"), ncol = 1,
                   rel_heights = 1:2, label_size = 18)
f5
```

```{r, include=FALSE}
save_plot("Figures/Figure5.pdf", plot = f5, base_width = 6.69, base_height = 9)
```
<br>
<br>



### 4. An example demonstrated the strong informativeness of covariate mean {#example}
***
```{r, fig.width=7, fig.height=2.5}
EWAS33 <- readRDS("Data/dataset/EWAS33.RDS")
candidate_covariates <- c("sd.b", "mad", "mean", "precision")

plot_data <- data.frame()
for(s in candidate_covariates){
  df <- data.frame(covariate_value = EWAS33[[s]],
                   p_value = EWAS33$P.value)
  df$covariate <- s
  f <- ecdf(df$covariate_value)
  df$percentile_covariate <- f(df$covariate_value)
  plot_data <- rbind(plot_data, df)
}

fs7 <- ggplot(data = plot_data, mapping = aes(x = percentile_covariate, y = -log10(p_value))) +
  geom_hex(bins = 100) + 
  facet_wrap(~covariate, nrow = 1) + 
  theme_bw() +
  theme(legend.key.size = unit(0.3, "cm"))
fs7
```

```{r, include=FALSE}
save_plot("Figures/FigureS7.pdf", plot = fs7, base_width = 7, base_height = 2.5)
```

